---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

install the packages which aren't installed in a new system

dplyr - all_equal

```{r}
#install.packages("dplyr")
```

load the packages

```{r}
library("tidyverse")
library("lubridate")
library("ggplot2")
library("dplyr")
```

set the working directory

```{r}
setwd("P:/Codes/E learning/Coursera/Professional Certificate/Google Data Analytics/8 Google Data Analytics Capstone Complete a Case Study/Week 2 Building your portfolio/Case study 1 Cyclistic")
```

STEP 1: COLLECT DATA


```{r}
nov_2020 <- read_csv("Dataset/Previous 12 months/CSV/202011-divvy-tripdata.csv")
dec_2020 <- read_csv("Dataset/Previous 12 months/CSV/202012-divvy-tripdata.csv")
jan_2021 <- read_csv("Dataset/Previous 12 months/CSV/202101-divvy-tripdata.csv")
feb_2021 <- read_csv("Dataset/Previous 12 months/CSV/202102-divvy-tripdata.csv")
mar_2021 <- read_csv("Dataset/Previous 12 months/CSV/202103-divvy-tripdata.csv")
apr_2021 <- read_csv("Dataset/Previous 12 months/CSV/202104-divvy-tripdata.csv")
may_2021 <- read_csv("Dataset/Previous 12 months/CSV/202105-divvy-tripdata.csv")
jun_2021 <- read_csv("Dataset/Previous 12 months/CSV/202106-divvy-tripdata.csv")
jul_2021 <- read_csv("Dataset/Previous 12 months/CSV/202107-divvy-tripdata.csv")
aug_2021 <- read_csv("Dataset/Previous 12 months/CSV/202108-divvy-tripdata.csv")
sep_2021 <- read_csv("Dataset/Previous 12 months/CSV/202109-divvy-tripdata.csv")
oct_2021 <- read_csv("Dataset/Previous 12 months/CSV/202110-divvy-tripdata.csv")
```

STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

check the structure of each tibbles

```{r}
str(nov_2020)
str(dec_2020)
str(jan_2021)
str(feb_2021)
str(mar_2021)
str(apr_2021)
str(may_2021)
str(jun_2021)
str(jul_2021)
str(aug_2021)
str(sep_2021)
str(oct_2021)
```

Convert start_station_id and end_station_id to character so that they can stack correctly

```{r}
nov_2020 <- mutate(nov_2020, start_station_id = as.character(start_station_id),
                   end_station_id = as.character(end_station_id))
```

Stack individual month's data frames into season (as per Chicago) wise data frame

winter - dec +jan + feb 
summer - jun + jul + aug
spring - mar + apr + may
autumn - sep + oct + nov
```{r}

```

Stack individual month's data frames into one big data frame for an year
```{r}
nov20_to_oct21 <- bind_rows(nov_2020,
                            dec_2020,
                            jan_2021,
                            feb_2021,
                            mar_2021,
                            apr_2021,
                            may_2021,
                            jun_2021,
                            jul_2021,
                            aug_2021,
                            sep_2021,
                            oct_2021)
```

Remove lat, long from nov20_to_oct21

```{r}
nov20_to_oct21 <- nov20_to_oct21 %>%
  select(-c(start_lat,start_lng,end_lat,end_lng))

```

STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

```{r}
head(nov20_to_oct21)
str(nov20_to_oct21)
summary(nov20_to_oct21)
```

Check how many observations fall under each user type

```{r}
table(nov20_to_oct21$member_casual)
```

Reassign to the desired values and recheck if the observations were reassigned

```{r}
nov20_to_oct21 <- nov20_to_oct21 %>%
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer" = "casual"))
table(nov20_to_oct21$member_casual)
```

Add columns that list the date, month, day, and year of each ride

```{r}
nov20_to_oct21$date <- as.Date(nov20_to_oct21$started_at)
nov20_to_oct21$year <- format(as.Date(nov20_to_oct21$started_at),"%Y")
nov20_to_oct21$month <- format(as.Date(nov20_to_oct21$started_at),"%m")
nov20_to_oct21$day <- format(as.Date(nov20_to_oct21$started_at),"%d")
nov20_to_oct21$day_of_week <- format(as.Date(nov20_to_oct21$started_at),"%A")
```

Reassign to the Month names and recheck if the observations were reassigned

```{r}
nov20_to_oct21 <- nov20_to_oct21 %>%
  mutate(month = recode(month,
                      "01" = "January",
                      "02" = "February",
                      "03" = "March",
                      "04" = "April",
                      "05" = "May",
                      "06" = "June",
                      "07" = "July",
                      "08" = "August",
                      "09" = "September",
                      "10" = "October",
                      "11" = "November",
                      "12" = "December"))
table(nov20_to_oct21$month)
```

Add ride_length column

```{r}
nov20_to_oct21$ride_length <- difftime(nov20_to_oct21$ended_at,nov20_to_oct21$started_at)
```

Inspecting the strucutre of the columns after adding the data as per the requirements

```{r}
str(nov20_to_oct21)

is.factor(nov20_to_oct21$ride_length)
nov20_to_oct21$ride_length <- as.numeric(as.character(nov20_to_oct21$ride_length))
is.numeric(nov20_to_oct21$ride_length)
summary(nov20_to_oct21$ride_length)
```

Remove Bad data

```{r}
nov20_to_oct21_v2 <- nov20_to_oct21[!(nov20_to_oct21$start_station_name == "HQ QR" 
                                      | nov20_to_oct21$ride_length < 0),]
```

Descriptive Analysis

```{r}
summary(nov20_to_oct21_v2$ride_length)
```

Compare members and casual riders w.r.t ride_length

```{r}
aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual, FUN=mean)
aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual, FUN=median)
aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual, FUN=min)
aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual, FUN=max)
```

Check the average ride time by each day for members vs casual users

```{r}
aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual + nov20_to_oct21_v2$day_of_week, FUN=mean)
```

Arranging the days of the week in order

```{r}
nov20_to_oct21$day_of_week <- ordered(nov20_to_oct21$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
nov20_to_oct21_v2$day_of_week <- ordered(nov20_to_oct21_v2$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

analyze ridership data by type and weekday

```{r}
nov20_to_oct21_v2 %>%
  mutate(weekday=wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```

visualize the number of rides by rider type

```{r}
nov20_to_oct21_v2 %>%
  mutate(weekday=wday(started_at, label=TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + 
  geom_col(position = "dodge")
```

Visualize the average duration by rider types

```{r}
nov20_to_oct21_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday) %>%
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
  
```
```{r}
counts <- aggregate(nov20_to_oct21_v2$ride_length ~ nov20_to_oct21_v2$member_casual 
                    + nov20_to_oct21_v2$day_of_week, FUN = mean)
write.csv(counts, file="P:/Codes/E learning/Coursera/Professional Certificate/Google Data Analytics/8 Google Data Analytics Capstone Complete a Case Study/Week 2 Building your portfolio/Case study 1 Cyclistic/Findings/avg_ride_length.csv")
```

